Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-08-05 analysis/bai/analysis_resilience_bai.Rmd
update review GGI

---
title: "Growth Resilience (reanalisis)"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Dec"
output: 
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

@@

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("tidyverse")
library("dplR")
library("knitr")
library('gtable')
library('grid')
library('gridExtra')
library('stringr')
library('pander')
library('broom')
library('effects')
library('devtools')
# devtools::install_github("ajpelu/auxiliar")
library('car')
library('auxiliar')
library('WRS2')
library('MASS')
library('rcompanion')
library('here')
```
# Resilience

@@

# Compute diameter (mm)
compete <- compete %>% 
  mutate(dn_mm = (perim_mm / pi))

>>>>

---
title: "Growth Resilience (reanalisis)"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Dec"
output: 
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
editor_options: 
  chunk_output_type: console
---

@@

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("tidyverse")
library("dplR")
library("knitr")
library('gtable')
library('grid')
library('gridExtra')
library('stringr')
library('pander')
library('broom')
library('effects')
library('devtools')
# devtools::install_github("ajpelu/auxiliar")
library('car')
#library('auxiliar')
library('WRS2')
library('MASS')
library('rcompanion')
library('here')
```
# Resilience

@@

# Compute diameter (mm)
compete <- compete %>% 
  dplyr::mutate(dn_mm = (perim_mm / base::pi))

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update review GGI

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge"
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

@@

## **Short Title**
Resilience of relict oaks in the rear-edge (max. 45 characters)

## Authors [^+^] 
A.J. Perez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^

@@

[^+^]: Author Contributions. AJPL, GGI and RZ conceived of or design the study. AJPL, GGI and RZ conducted field work and collected the data. AJPL and GGI conducted lab work and led article preparation. AJPL analyzed data and wrote the paper. GGI and RZ contributed in the writing process. All authors contribute to the drafts and gave final approval for publication.

## Manuscript highlights
* Forest resilience of *Quercus pyrenaica* rear-edge forests were analyzed at tree and stand-level
* Response to climate and resilience following a water-stress gradient
* Trees and stand expressed a high sensitivity to drought and to land-use legacies

## Abstract
* Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified the forest structure as result of intensive human-pressures in the past. We analyzed forest resilience of adult trees to drought and climate change of relict Mediterranean *Quercus pyrenaica* Willd. populations in the Southern Iberian Peninsula. Forest resilience was analyzed using both dendroecological growth estimates and greenness (NDVI) as proxies for secondary and primary growth respectively. Additionally forest history were inferred from an exhaustive review of historical documents.

* Intense land-use were likely expressed by disturbance chronologies, reflecting the origin of the current forest structure and land-use legacies. Trees expressed a high sensitivity to drought. Resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and to long-term climatic changes. Response to climate and resilience following a water-stress gradient. We found a positive growth trend since the end of the 1970s, that was strongest in the high-elevation stand and non-significant in the most xeric site. Northern populations were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.

>>>>

---
title: "Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies "
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

@@

## **Short Title**
Relict oaks resilience at their rear-edge

## Authors [^\*^] 
A.J. Perez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^

@@

[^\*^]: Author Contributions. AJPL, GGI and RZ conceived the study, conducted field work and collected the data. AJPL and GGI conducted lab work and led article preparation. AJPL analyzed data and led the writing of the paper. GGI and RZ contributed in the writing process. All authors contributed to the drafts and gave final approval for publication.

## Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
* The responses to climate and resilience followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

## Abstract

* Global change challenges ecosystems in dry locations transformed by intensive human use. We analyzed resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd. populations in the Southern Iberian Peninsula. Forest resilience to drought was analyzed in relation to historical records of land-use using dendroecological growth and greenness (EVI) as proxies for secondary and primary growth.

* The response to climate and resilience followed a water-stress gradient. Primary and secondary growth of more xeric populations (*i.e.* Northern aspect and low elevations) was less resilient to drought than that at more mesic sites. In addition, a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic high-elevation stands, but absent in the most xeric site. Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term. The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.

* The species showed high resilience to drought despite being relict and at its rear-edge. Yet, the trees exhibited a variable response to the changing climate along a very short climatic gradient. The trees were more sensitive to water availability at most xeric sites whereas only the coldest and more humid stands exhibited a recent positive growth trend. Stand dynamics at the most xeric locations (*i.e.* the altitudinal rear-edge) needs to be monitored under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.

### 2019-08-05 man/tables/t_dendro_summary.Rmd
update review GGI

**Table 2.** Characteristics of the mean tree ring chronologies. Values of the length year in parenthesis indicate years replicated with more than five series. RW = mean annual ring width (standard deviation in parenthesis). MS = mean sensitivity. AR(1) = mean autocorrelation of raw series. Rbt = mean correlation between series. EPS = mean expressed population signal. EPS and Rbt are calculated for the mean residual chronologies of growth indices.

>>>>

**Table 2.** Characteristics of the mean tree ring chronologies. Values of length between parentheses indicate number of years replicated with more than five series. RW = mean annual ring width (standard deviation in parenthesis). MS = mean sensitivity. AR(1) = mean autocorrelation of raw series. Rbt = mean correlation between series. EPS = mean expressed population signal. EPS and Rbt were calculated for the mean residual chronologies of growth indices.

### 2019-08-05 man/tables/t_huber_res_bai.Rmd
update review GGI

**Table S2.** Robust measures of central tendency of resilience indices for tree-growth (BAI) grouped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included in parentheses.

@@

mt <- regulartable(mi, col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
                                "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
mt <- mt %>% 
  # Header 
  set_header_labels(site = "Sites", 
                        '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
                        '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience") %>% 
  add_header(site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
                        '2099rt' = "", '2099rc' = "", '2099rs' = "", top = TRUE) %>% 
  merge_at(i = 1, j =c(3:5) , part = "header") %>% 
  merge_at(i = 1, j =c(7:9) , part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 9) %>%

>>>>

**Table S2.** Robust measures of central tendency of resilience indices for tree-growth (BAI) grouped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included between parentheses.

@@

mt <- regulartable(mi, col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
                                "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
mt <- mt %>% 
  # Header 
  set_header_labels(site = "Sites", 
                        '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
                        '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience") %>% 
  add_header(site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
                        '2099rt' = "Total", '2099rc' = "Total", '2099rs' = "Total", top = TRUE) %>% 
  merge_at(i = 1, j =c(3:5) , part = "header") %>% 
  merge_at(i = 1, j =c(7:9) , part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 9) %>%

### 2019-08-05 man/tables/t_huber_res_evi.Rmd
update review GGI

**Table S1.** Robust measures of central tendency of resilience indices for greenness (EVI) grouped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included in parentheses.

@@

mt <- regulartable(mi, col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
                                "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
mt <- mt %>% 
  # Header 
  set_header_labels(site = "Oak populations", 
                        '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
                        '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience") %>% 
  add_header(site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
                        '2099rt' = "", '2099rc' = "", '2099rs' = "", top = TRUE) %>% 
  merge_at(i = 1, j =c(3:5) , part = "header") %>% 
  merge_at(i = 1, j =c(7:9) , part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 9) %>%

>>>>

**Table S1.** Robust measures of central tendency of resilience indices for greenness (EVI) grouped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included between parentheses.

@@

mt <- regulartable(mi, col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
                                "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
mt <- mt %>% 
  # Header 
  set_header_labels(site = "Oak populations", 
                        '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
                        '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience") %>% 
  add_header(site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
                        '2099rt' = "Total", '2099rc' = "Total", '2099rs' = "Total", top = TRUE) %>% 
  merge_at(i = 1, j =c(3:5) , part = "header") %>% 
  merge_at(i = 1, j =c(7:9) , part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 9) %>%

### 2019-08-05 man/tables/t_robust_anovas.Rmd
update review GGI

mti <- mt %>% 
  autofit() %>% 
  empty_blanks() %>% 
  align(j = 1,  align = "center") %>% 
  merge_v(~variable) %>% 
  set_header_labels(variable = "", factores = "", rt_F = "Resistance", rt_p = "Resistance", rc_F = "Recovery", rc_p = "Recovery", rs_F = "Resilience", rs_p = "Resilience") %>% 
  add_header(variable = "Variable", factores = "Factor", rt_F = "F", rt_p = "p", rc_F = "F", rc_p = "p", rs_F = "F", rs_p = "p", top = FALSE) %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 10) %>%
  fontsize(part = "body", size = 9) %>% 
  width(j = 2, width = 2) %>% 
  italic(i = ~ rt_p < 0.01, j = ~ rt_p) %>% 
  italic(i = ~ rc_p < 0.01, j = ~ rc_p) %>% 
  italic(i = ~ rs_p < 0.01, j = ~ rs_p) %>% 
  merge_at(i = 1, j=4:5, part="header") %>% 
  merge_at(i = 1, j=7:8, part="header") %>% 
  merge_at(i = 1, j=10:11, part="header")

>>>>

mti <- mt %>% 
  autofit() %>% 
  empty_blanks() %>% 
  align(j = 1,  align = "center") %>% 
  merge_v(~variable) %>% 
  set_header_labels(variable = "", factores = "", rt_F = "Resistance", rt_p = "Resistance", rc_F = "Recovery", rc_p = "Recovery", rs_F = "Resilience", rs_p = "Resilience") %>% 
  add_header(variable = "Variable", factores = "Factor", rt_F = "F", rt_p = "p", rc_F = "F", rc_p = "p", rs_F = "F", rs_p = "p", top = FALSE) %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 10) %>%
  fontsize(part = "body", size = 9) %>% 
  width(j = 2, width = 2) %>% 
  italic(i = ~ rt_p < 0.01, j = ~ rt_p) %>% 
  italic(i = ~ rc_p < 0.01, j = ~ rc_p) %>% 
  italic(i = ~ rs_p < 0.01, j = ~ rs_p) %>% 
  compose(i = ~rt_p == 0.001, j = ~rt_p,
          value = as_paragraph("< ", as_chunk(rt_p))) %>% 
  compose(i = ~rc_p == 0.001, j = ~rc_p,
          value = as_paragraph("< ", as_chunk(rc_p))) %>% 
  compose(i = ~rs_p == 0.001, j = ~rs_p,
          value = as_paragraph("< ", as_chunk(rs_p))) %>% 
  merge_at(i = 1, j=4:5, part="header") %>% 
  merge_at(i = 1, j=7:8, part="header") %>% 
  merge_at(i = 1, j=10:11, part="header")

### 2019-08-05 man/tables/t_sampled_plots.Rmd
update review GGI

**Table 1.** Characteristics of sampled plot. Lat = latitude; Long = longitude. Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods). Temp.: annual average of mean monthly minimun and maximum temperatures. Values shown here correspond to site averages. Standard deviations are shown in parentheses. Different letters indicate statistically significant differences between sites (Kruskal-Wallis test followed by Dunn’s test, p<0.05).

@@

ta <- ta %>% set_header_labels(name = "Site", lat_m = "Lat (\u00B0)", long_m = "Long (\u00B0)", 
                               elev = "Elevation\n(m)", slope = "Slope (\u00B0)",  prec = "Prec.\n(mm)",
                               temp = "Temp.\n(\u00B0 C)", n_trees = "# trees\n(# cores)", dn = "Dbh (cm)",
                               h = "Height\n(m)", age = "Age\n(years)", dn_all = "Dbh all\n(cm)", 
                               h_all = "Height all\n(m)", ba = "BA\n(m2/ha)", 
                               stand_density = "Density\n(trees/ha)", srd = "SRD") %>% 
  add_header(name="", lat_m ="", long_m="", elev="", slope ="", prec = "", temp = "",
             n_trees = "Cored trees", dn = "Cored trees", h = "Cored trees", age = "Cored trees", 
             dn_all = "Competition", h_all = "Competition", ba = "Competition", 
             stand_density = "Competition", srd = "Competition", top = TRUE) %>% 
  merge_at(i = 1, j =c(8:11) , part = "header") %>% 
  merge_at(i = 1, j =c(12:16) , part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(i= 1, part = "header", size = 8) %>% 
  fontsize(i= 2, part = "header", size = 8) %>% 
  italic(i=1, part = "header") %>% 
    # General format 
  fontsize(part = "body", size = 8) %>% 
  autofit(add_w = 0) %>% 
  align(align = "center") %>% 
  width(j = c(2,3), width = 0.35) %>% 
  width(j = c(6,7,8), width = 0.6)

>>>>

**Table 1.** Characteristics of sampled plot. Lat = latitude; Long = longitude. Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods). Temp.: annual average of mean monthly minimun and maximum temperatures. Values shown here correspond to site averages. Standard deviations are shown in parentheses. Different letters indicate statistically significant differences between sites (Kruskal-Wallis test followed by Dunn’s test, p<0.05). Stands were monospecific, hence all results correspond to oak data.

@@

ta <- ta %>% set_header_labels(name = "Site", lat_m = "Lat (\u00B0)", long_m = "Long (\u00B0)", 
                               elev = "Elevation\n(m)", slope = "Slope (\u00B0)",  prec = "Prec.\n(mm)",
                               temp = "Temp.\n(\u00B0 C)", n_trees = "# trees\n(# cores)", dn = "Dbh (cm)",
                               h = "Height\n(m)", age = "Age\n(years)", dn_all = "Dbh all\n(cm)", 
                               h_all = "Height all\n(m)", ba = "BA\n(m2/ha)", 
                               stand_density = "Density\n(trees/ha)", srd = "SRD") %>% 
  add_header(name="", lat_m ="", long_m="", elev="", slope ="", prec = "", temp = "",
             n_trees = "Cored trees", dn = "Cored trees", h = "Cored trees", age = "Cored trees", 
             dn_all = "Stand competition", h_all = "Stand competition", ba = "Stand competition", 
             stand_density = "Stand competition", srd = "Stand competition", top = TRUE) %>% 
  merge_at(i = 1, j =c(8:11) , part = "header") %>% 
  merge_at(i = 1, j =c(12:16) , part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(i= 1, part = "header", size = 8) %>% 
  fontsize(i= 2, part = "header", size = 8) %>% 
  italic(i=1, part = "header") %>% 
    # General format 
  fontsize(part = "body", size = 8) %>% 
  autofit(add_w = 0) %>% 
  align(align = "center") %>% 
  width(j = c(2,3), width = 0.35) %>% 
  width(j = c(6,7,8), width = 0.6)
